Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-01-05 sources/bmc_template_submission/bmc_article.tex
intro again

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations.
Public biological databases constitute an amazing resource for evolutionary
studies. Updating existing phylogenies with molecular data that has never been
incorporated into any phylogenetic estimate, geographical location, fossils, and
other data in a reproducible and continuous manner is possible by establishing a
data interoperability framework for biological databases. Here, we introduce
Physcraper, a tool that automates database connections to build upon homology
hypotheses that taxon specialists have assessed and deemed appropriate for a
specific phylogenetic scope to update a starting tree and single locus alignments
with public DNA data.

@@

Another challenge for incorporating public molecular data into existing phylogenies
is assembling high-quality homology hypotheses.
While genomics has, and will continue to, revolutionize phylogenetic inference,
the variety of alternative genomic sequencing approaches it uses produce largely
non-overlapping genomic datasets across taxa, creating challenges in wide scale
phylogenetic reconstruction.

Phylogenomics ameliorate this problem by focusing on targeted capture of informative
loci \cite{andermann2020guide}. Yet, decades of single locus sequencing have generated
massive amounts of homologous DNA datasets that can be used for phylogenetic
reconstruction at many scales.

More than a decade ago, GenBank release 159 (April 15, 2007) already hosted 72
million DNA sequences that were gauged to have the potential to resolve phylogenetic
relationships of 98.05\% of the almost 241,000
distinct taxa in the NCBI taxonomy at the time \cite{sanderson2008phylota}.
Assembling a DNA alignment from such a massive database can be done "by hand",
but it is a largely time consuming and mostly non-reproducible approach.
Computational pipelines that mine DNA databases fast, efficiently, and reproducibly,
have been applied to infer phylogenetic relationships in a variety of organisms
(e.g., \cite{smith2009mega, antonelli2017toward, izquierdo2014pumper}).
However, fine-grained curated markers and alignments can improve phylogenetic
reconstructions, even in phylogenomic analyses \cite{fragoso2017pilot}.

>>>>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{Dobzhansky maybe}.
A challenge for phylogenetic reconstruction
is assembling high-quality homology hypotheses.
While genomics has, and will continue to, revolutionize phylogenetic inference,
the diversity of alternative genomic sequencing approaches implemented produce
largely non-overlapping homology hypotheses across taxa \cite{BLA}, creating challenges for
reconstructing the shared evolutionary history of taxa separated phylogenetically
at deep scales \cite{BLA}.

While phylogenomics tackles this problem by focusing on targeted capture of informative
regions \cite{andermann2020guide, it is expensive. Few labs around the world can afford
to implement it for all their study groups, or at all. It does not contribute to
democratization of research yet.

Decades of single locus sequencing have generated
massive amounts of homologous DNA datasets that can be used for phylogenetic
reconstruction at many scales.
More than a decade ago, GenBank release 159 (April 15, 2007) already hosted 72
million DNA sequences that were gauged to have the potential to resolve phylogenetic
relationships of 98.05\% of the almost 241,000
distinct taxa in the NCBI taxonomy at the time \cite{sanderson2008phylota}.
How much of this is available on OpenTree??

Is genomic data on genbank available for a smaller percentage of all life, compared to
single loci in the same database?

Notable examples of phylogenies constructed using multi-locus, database mined, super matrix approach:
Magallon, Soltis, examples with fish, bats,

These studies are widely cited and immensely relevant for the biological community.

Some of these multilocus phylogenetic studies have implemented pipelines such as PHLAWD,
other studies have curated their matrices by hand, which is largely non-reproducible.

Public biological databases constitute an amazing resource for evolutionary
studies. Updating existing phylogenies with molecular data that has never been
incorporated into any phylogenetic estimate, geographical location, fossils, and
other data in a reproducible and continuous manner is possible by establishing a
data interoperability framework for biological databases. Here, we introduce
Physcraper, a tool that automates database connections to build upon homology
hypotheses that taxon specialists have assessed and deemed appropriate for a
specific phylogenetic scope to update a starting tree and single locus alignments
with public DNA data.

@@

----------

Assembling a DNA alignment from such a massive database can be done "by hand",
but it is a largely time consuming and mostly non-reproducible approach.
Computational pipelines that mine DNA databases fast, efficiently, and reproducibly,
have been applied to infer phylogenetic relationships in a variety of organisms
(e.g., \cite{smith2009mega, antonelli2017toward, izquierdo2014pumper}).
However, fine-grained curated markers and alignments can improve phylogenetic
reconstructions, even in phylogenomic analyses \cite{fragoso2017pilot}.
